Notes on Perturbational Quantities

### 2011-06-15 docs/pert_notes/pert.tex
finish species_pred_type stuff (for what's in the code now)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Advancing $\rho X_k$}\label{Advancing rhoX_k}
The evolution equation for $\rho X_k$, ignoring the reaction terms
that were already accounted for in {\tt react\_state}, and the
associated discretization is
\begin{equation}
\frac{\partial\rho X_k}{\partial t} = -\nabla\cdot(\rho X_k\Ub) = 
-\nabla\cdot\left\{\left[\left(\rho_0^{n+\myhalf,{\rm avg}} 
+ \rho'\right)X_k\right](\Ubt+w_0\eb_r)\right\}.
\end{equation}

%-----------------------------------------------------------------------------
% energy
%-----------------------------------------------------------------------------
\section{Energy Evolution}
We use {\tt enthalpy\_pred\_type} = 1, in which we predict $(\rho h)'$ at edges.
The full enthalpy equation is
\begin{eqnarray}
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h \Ub) + \frac{Dp_0}{Dt} 
+ \nabla\cdot \kth \nabla T + \rho H_{\rm nuc} + \rho H_{\rm ext} \nonumber \\
&=& \underbrace{-\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub}_{-\nabla\cdot(\rho h\Ub)} 
+ \underbrace{\psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r}}_{\frac{Dp_0}{Dt}} 
+ \nabla\cdot\kth\nabla T + \rho H_{\rm nuc} + \rho H_{\rm ext}.
\end{eqnarray}
%The full temperature equation is
%\begin{equation}
%\frac{\partial T}{\partial t} = -\Ub\cdot\nabla T
%+ \frac{1}{\rho c_p}\left\{(1-\rho h_p)\left[\psi
%+ (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}\right] - \sum_k\rho\xi_k\dot\omega_k 
%+ \rho H_{\rm nuc} + \rho H_{\rm ext}\right\}.
%\end{equation}
Due to operator splitting, the call to {\tt react\_state} has already
been made.  Hence, the goal is to compute an edge state enthalpy
starting from $(\rho h)^{(1)}$ using an enthalpy equation that does
not include the $\rho H_{\rm nuc}$ and $\rho H_{\rm ext}$ terms, where
were already accounted for in {\tt react\_state}, so our equation becomes
\begin{equation}
\frac{\partial(\rho h)}{\partial t} = -\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub 
+ \psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} + \nabla\cdot\kth\nabla T. \label{rhoh equation}
\end{equation}
We define the base state enthalpy evolution equation as
\begin{eqnarray}
\frac{\partial(\rho h)_0}{\partial t} &=& -\nabla\cdot[(\rho h)_0 w_0\eb_r] 
+ \frac{D_0p_0}{Dt} \nonumber \\
&=& -w_0\frac{\partial(\rho h)_0}{\partial r} 
- \underbrace{(\rho h)_0\frac{\partial w_0}{\partial r}+ \psi}_{"(\rho h)_0 ~ \text{force}"}
\enskip .\label{rhoh0 equation}
\end{eqnarray}
Subtracting (\ref{rhoh0 equation}) from (\ref{rhoh equation}) and rearranging terms gives 
the perturbational enthalpy equation
\begin{eqnarray}
\frac{\partial(\rho h)'}{\partial t} &=& -\nabla\cdot[(\rho h)'\Ub] 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r} 
+ \nabla\cdot\kth\nabla T\nonumber \\
&=& -\Ub\cdot\nabla(\rho h)' \underbrace{- (\rho h)'\nabla\cdot\Ub 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r}
+ \nabla\cdot\kth\nabla T}_{"(\rho h)' ~ \text{force}"}, \label{rhohprime equation}
\end{eqnarray}
where the first two terms in $(\rho h)'$ force are computed in 
{\tt modify\_scal\_force}, and the last two terms are accounted for in
{\tt mkrhohforce}.  For spherical problems, we have found that a different 
representation of the pressure term in the $(\rho h)'$ force gives better
results, namely:
\begin{equation}
(\Ubt \cdot \er)\frac{\partial p_0}{\partial r} \equiv \Ubt\cdot\nabla p_0 = 
\nabla\cdot(\Ubt p_0) - p_0\nabla\cdot\Ubt.
\end{equation}

>>>>

We note that even though it is not needed here, we still predict
$\rho'$ to the edges since certain enthalpy formulations need it.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Advancing $\rho X_k$}\label{Advancing rhoX_k}
The evolution equation for $\rho X_k$, ignoring the reaction terms
that were already accounted for in {\tt react\_state}, and the
associated discretization is \\

\noindent {\tt species\_pred\_type = pred\_rhoprime\_and\_X}:
\begin{equation}
\frac{\partial\rho X_k}{\partial t} = -\nabla\cdot(\rho X_k\Ub) = 
-\nabla\cdot\left\{\left[\left({\rho_0}^{n+\myhalf,{\rm avg}}
+ \rho'_\mathrm{edge} \right)(X_k)_\mathrm{edge} \right](\Ubt+w_0\eb_r)\right\}.
\end{equation} \\

\noindent {\tt species\_pred\_type = pred\_rhoX}:
\begin{equation}
\frac{\partial\rho X_k}{\partial t} = -\nabla\cdot(\rho X_k\Ub) = 
-\nabla\cdot\left\{\left[\left(\rho X_k \right)_\mathrm{edge} \right](\Ubt+w_0\eb_r)\right\}.
\end{equation}

%-----------------------------------------------------------------------------
% energy
%-----------------------------------------------------------------------------
\section{Energy Evolution}

Note: only the default {\tt enthalpy\_pred\_type = 1} is guaranteed to work
with {\tt species\_pred\_type > 1}. \\

We use {\tt enthalpy\_pred\_type} = 1, in which we predict $(\rho h)'$ at edges.
The full enthalpy equation is
\begin{eqnarray}
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h \Ub) + \frac{Dp_0}{Dt} 
+ \nabla\cdot \kth \nabla T + \rho H_{\rm nuc} + \rho H_{\rm ext} \nonumber \\
&=& \underbrace{-\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub}_{-\nabla\cdot(\rho h\Ub)} 
+ \underbrace{\psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r}}_{\frac{Dp_0}{Dt}} 
+ \nabla\cdot\kth\nabla T + \rho H_{\rm nuc} + \rho H_{\rm ext}.
\end{eqnarray}
%The full temperature equation is
%\begin{equation}
%\frac{\partial T}{\partial t} = -\Ub\cdot\nabla T
%+ \frac{1}{\rho c_p}\left\{(1-\rho h_p)\left[\psi
%+ (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}\right] - \sum_k\rho\xi_k\dot\omega_k 
%+ \rho H_{\rm nuc} + \rho H_{\rm ext}\right\}.
%\end{equation}
Due to operator splitting, the call to {\tt react\_state} has already
been made.  Hence, the goal is to compute an edge state enthalpy
starting from $(\rho h)^{(1)}$ using an enthalpy equation that does
not include the $\rho H_{\rm nuc}$ and $\rho H_{\rm ext}$ terms, where
were already accounted for in {\tt react\_state}, so our equation becomes
\begin{equation}
\frac{\partial(\rho h)}{\partial t} = -\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub 
+ \psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} + \nabla\cdot\kth\nabla T. \label{rhoh equation}
\end{equation}
We define the base state enthalpy evolution equation as
\begin{eqnarray}
\frac{\partial(\rho h)_0}{\partial t} &=& -\nabla\cdot[(\rho h)_0 w_0\eb_r] 
+ \frac{D_0p_0}{Dt} \nonumber \\
&=& -w_0\frac{\partial(\rho h)_0}{\partial r} 
- \underbrace{(\rho h)_0\frac{\partial w_0}{\partial r}+ \psi}_{"(\rho h)_0 ~ \text{force}"}
\enskip .\label{rhoh0 equation}
\end{eqnarray}
Subtracting (\ref{rhoh0 equation}) from (\ref{rhoh equation}) and rearranging terms gives 
the perturbational enthalpy equation
\begin{eqnarray}
\frac{\partial(\rho h)'}{\partial t} &=& -\nabla\cdot[(\rho h)'\Ub] 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r} 
+ \nabla\cdot\kth\nabla T\nonumber \\
&=& -\Ub\cdot\nabla(\rho h)' \underbrace{- (\rho h)'\nabla\cdot\Ub 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r}
+ \nabla\cdot\kth\nabla T}_{"(\rho h)' ~ \text{force}"}, \label{rhohprime equation}
\end{eqnarray}
where the first two terms in $(\rho h)'$ force are computed in 
{\tt modify\_scal\_force}, and the last two terms are accounted for in
{\tt mkrhohforce}.  For spherical problems, we have found that a different 
representation of the pressure term in the $(\rho h)'$ force gives better
results, namely:
\begin{equation}
(\Ubt \cdot \er)\frac{\partial p_0}{\partial r} \equiv \Ubt\cdot\nabla p_0 = 
\nabla\cdot(\Ubt p_0) - p_0\nabla\cdot\Ubt.
\end{equation}
